Draft -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- Draft

### 2024-10-14 docs/_tex/index.tex
adding bede's comments

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing can be limited by too coarse spatial
and/or spectral resolution, making it difficult to discriminate seagrass
from other macrophytes in highly heterogenous meadows. Drone (unmanned
aerial vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor similar to that onboard
Sentinel-2, for mapping intertidal macrophytes at low tide (i.e.~during
a period of emersion) and effectively discriminating between seagrass
and green macroalgae. Nine drone flights were conducted at two different
altitudes (12 m and 120 m) across heterogeneous intertidal European
habitats in France and Portugal, providing multispectral reflectance
observation at very high spatial resolution (respectively at 8 mm and 80
mm). Taking advantage of their extremly high spatial resolution; the low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discrimintate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms), and validated using concomitant
field measurements. Classification of drone imagery resulted in an
overall accuracy of 94\% across all sites and images, covering a total
area of 467 000 m². The model exhibited an accuracy of 96.4\% in
identifying seagrass. In particular , seagrass and green algae can be
discriminated. The very high spatial resolution of the drone data made
ot possoble to assess the influence of spatial resolution on the
classification outputs, showing a limited loss in seagrass detection up
to about 10m. Altogether, out findings suggest that the Multi-Spectral
Instrument (MSI) onboard Sentinel-2 offers a relevant trade-off between
its spatial and spectral resolution, thus offering promising
perspectives for satellite remote sensing of intertidal biodiversity
over lager scales.
\end{abstract}

@@

The importance of seagrass meadows and the variety of ecosystem services
they provide have led to the enhancement of both global and regional
programs to monitor Essential Oceanic Variable (EOVs) such as seagrass
composition \citep{Miloslavich2018}, as well as Essential Biodiversity
Variable (EBVs) such as seagrass taxonomic diversity, species
distribution, population abundance, and phenology \citep{Pereira2013}.
Traditionally, indicators of seagrass status have been quantified using
\emph{in situ} measurements. However, the acquisition of field
measurements in intertidal zones is notoriously challenging. Intertidal
seagrass meadows are only exposed during low tide and can be situated in
difficult-to-reach mudflats, potentially leading to inaccurate and
limited estimations with conventional sampling techniques
\citep{nijland2019}. Satellite observations have been proven effective
in complementing \emph{in situ} sampling, allowing for near real-time
and consistent retrieval of seagrass EOVs and EBVs over extensive
meadows \citep{Zoffoli2021, xu2021, Traganos2018, coffer2023}.

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a too
coarse spatial resolution (\textgreater100 m) to accurately map patchy
seagrass meadows. Missions with a high spatial resolution such as
Sentinel-2 (10 m) or Landsat8/9 (30 m) can be limited by low spectral
resolution. The limited number of spectral bands challenges accurate
discrimination of seagrass from other co-existing macrophytes. In
particular, Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition
\citep{ralph2002, Douay2022}, resulting in a similar spectral signature
in therms of reflectance, especially in the visible range
\citep{Davies2023, bannari2022}. Recently, using advanced
machine-learning algorithms trained with a large hyperspectral library
of more than 300 field reflectance spectra, \citep{Davies2023}
demonstrated that it was possible to discriminate Magnoliopsida from
Chlorophyceae using reflectance spectra at Sentinel-2 's spectral
resolution. However the application of this approach to satellite RS
remains to be validated. Moreover patches of green algae can develop at
small spatial scales that are not observable using Sentinel-2 and/or
Landsat-8/9 images \citep{tuya2013}, especially during the initial stage
of a green tide.

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management
\citep{adade2021, casella2020, angnuureng2022} to mapping species
distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue}.
However, when applied to coastal habitat mapping, previous case studies
were mostly limited to a limited number of flights over a single study
site, restricting the generalizability of their application over wider
geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
The present study aimed at analyzing the potential of multispectral
drone RS to map intertidal macrophytes over a diverse biogeographical
range, with a particular focus on discriminating Magnoliopsida and
Chlorophyceae (Seagrass and Green Algae, respectively). Nine drone
flights were performed over soft-bottom intertidal areas along the
Atlantic coastlines of two European countries (France and Portugal),
covering a wide range of habitats, from monospecific seagrass meadows to
meadows mixed with green, or red macroalgae. A deep learning algorithm
was trained and validated for macrophyte discrimination, emphasizing
applicability across diverse sites without a loss of prediction
accuracy. The classification maps obtained at a very high spatial
resolution with the drone were spatially degraded to satellite
resolutions, making it possible to asses the effect of spatial
resolution on classification accuracy; and provide insights for coastal
habitat mapping using satellite remote sensing.

@@

Seven study sites distributed between France and Portugal were selected
for their extensive intertidal seagrass beds. Two sites were located in
the Gulf of Morbihan (Figure~\ref{fig-map} A), France (47.5791°N,
2.8018°W). This gulf covers an area of 115 km² and is only connected to
the sea through a 900 m wide channel. A total of 53 small islands are
scattered across the gulf leading to 250 km of shorelines. Patchy
seagrass meadows can be found on many of these islands. One of the sites
within the gulf was on one its islands (Arz) and the other was located
further south on a mainland beach area (Duer). The Gulf of Morbihan is a
Natura 2000 site and a Regional Protected Area due to its rich
biodiversity, including its seagrass meadows, and is also classified as
a RAMSAR site, which highlights its significance as a wetland of
international importance. Two other sites were located in Bourgneuf Bay,
France (46.9849°N, 2.1488°W). This is a 340 km² semi-enclosed macrotidal
bay, protected from waves by Noirmoutier Island. Bourgneuf bay hosts a
large intertidal seagrass meadow of about 6 km²
\citep{ZOFFOLI2020112020}. Within this meadow, the sites observed by
drones (L'Epine and Barbatre, Figure~\ref{fig-map} B) contained
monospecific beds of \emph{Nanozostera noltei} (dwarf eelgrass, syn.
\emph{Zostera noltei}) with very little mixing with other macrophytes.
Bourgneuf Bay is also part of the Natura 2000 network and serves as a
RAMSAR site due to its critical habitat for migratory bird species and
its extensive seagrass meadows. Three sites were surveyed in the Ria de
Aveiro Coastal Lagoon in Portugal (40.6887°N, 8.6810°W). The extent of
this lagoon is \textasciitilde83 km² (at low tide) with many narrow
channels, large salt marshes and many mudflats that uncover at low tide
\citep{sousa2017blue}. It is connected to the open sea through a single
channel, with a tidal lag between the North and the South of the lagoon.
The southernmost site (Gafanha) is a mudflat located in the Mira channel
(one of the four main channels of the lagoon) whereas the two other
sites (Mataducos and Marinha Lanzarote) were situated in the middle of
the lagoon and only accessible by boat (Figure~\ref{fig-map} C). These
Portuguese sites are characterized by a more diverse intertidal
vegetation, where patches of seagrass intermingle with red, brown, and
green macroalgae. The Aveiro Lagoon, like the other study areas, is a
Natura 2000 site and a RAMSAR wetland, recognized for its rich mosaic of
habitats and importance for biodiversity, including migratory bird
species and intertidal vegetation.

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m² ,which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation : Bacillariophyceae
(unicellular benthic diatoms forming biofilms at the sediment surface
during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf
eelgrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red
macroalgae) (Figure~\ref{fig-vegetation}). Only homogeneous vegetation
patches extending over several meters were selected as ground control
points. Pictures of each quadrat were uploaded online to the Global
Biodiversity Information Facility (GBIF) platform \citep{BedeGbif}, a
public open portal to store and share biodiversity data. Each photograph
was also processed to estimate the percent cover of each type of
vegetation using an image processing software
\citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to error reduction in photo interpretations.

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft} was used to process images to obtain
multispectral orthomosaics of each flight. The process for
orthomosaicking was identical for every flight. First, tying key points
were detected inside of each image and between overlapping images in
order to obtain a sparse point cloud. This cloud was cleaned using
reprojection accuracy metric in order to remove noisy points. A dense
point cloud was then produced using a structure from motion algorithm. A
surface interpolation of this dense point cloud was made to obtain a
digital surface model (DSM), used to reconstruct the multispectral
ortho-image \citep{nebel2020review}. Low altitude drone flights produced
ortho-images with a very high spatial resolution (8 mm per pixel),
making it efficient to visually distinguish between the various types of
vegetation. High altitude flights allowed to cover larger areas and
produced images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

A dataset containing photo-interpreted drone reflectance pixels was
built to train a Neural Network model. The training pixels were
categorized into seven different classes, representing the various
habitats encountered at the different study sites: Sediment, Water,
Chlorophyceae, Magnoliopsida, Bacillariophyceae, Phaeophyceae and
Rhodophyceae. Only low-altitude flights (Table~\ref{tbl-flights}) were
used for training because their 8 mm spatial resolution allowed to avoid
spectral sub-pixel mixing and to accurately identify vegetation classes.
More than 418,000 pixels at 8 mm resolution from the 3 training flights
were used to train the model (Table~\ref{tbl-validationPX}). Twenty one
variables were used by the model as predictors: the ten raw spectral
bands of the Micasense RedEdge Dual MX multispectral camera (ranging
from 444 nm to 840 nm), the same ten spectral bands standardized using a
min/max transformation (Equation~\ref{eq-std} ; \citep{Cao2017}) and the
Normalized difference vegetation index (NDVI, Equation~\ref{eq-ndvi}).
Standardisation of spectral bands is commonly used to eliminate the
scaling differences between spectra and to limit the effect of biomass
on the spectra shape \citep{Douay2022, Davies2023}.

@@

A neural network classification model was built using the fastai
workflow \citep{howard2018fastai}. This model is composed by 2 hidden
layers and have a total of 26 054 trainable parameters. Parameters have
been fine tuned using 12 epoch to minimize the error rate.

@@

The workflow of this study revolves around two distinct flight heights
(12 and 120 m, Figure~\ref{fig-workflow}) where ensuring consistency
between reflectances at both heights is crucial. This comparison was
conducted at sites where low and high altitude flights overlapped. The
low altitude flights were resampled to the same spatial resolution and
grid as the high flights using a median resampling method. Reflectance
values were then extracted, and a scatterplot was generated.
Subsequently, the slope of the linear model and the coefficient of
determination (R²) were computed.

The model was applied to all flights at both 12 and 120 m of altitude.
\emph{In situ} information on georeferenced class type and percent
cover, acquired at over homogeneous vegetation patches at the same time
as drone flights was used to assess the model accuracy. These images
were used to construct a validation dataset indicating the presence or
absence of each class. Additionally to the quadrat-based validation
dataset, polygons of each class were photo interpreted in order to
increase the number of pixels of the validation dataset.A total of
536,000 pixels were used to validate the Neural Network classifier. The
sites with the lowest and highest number of validation data were Gafanha
Low (17316 pixels) and Marinha Lanzarote (159713 pixels), respectively.
A confusion matrix, along with precision metrics such as global
accuracy, sensitivity, specificity, and Kappa coefficient, was generated
for each sites. All validation matrices were then aggregated to create
an overall matrix .

@@

Variable Importance Plots (VIP) serve as a method to identify which
predictors are important for predicting a specific class. Out of the
twenty one predictors utilized in this study, Variable Importance was
computed only for the raw and standardized values of the 10 spectral
bands captured by the MicaSense camera. This is achieved by repeatedly
predicting the same dataset while randomly shuffling one predictor at a
time. The benchmark score obtained after each iteration is then compared
to the benchmark score obtained without shuffling any variables. The
greater the difference between these two benchmark values, the more
important the variable is for the model \citep{WEI2015399}.

@@

We used a Generalized Linear Model (GLM) with a Gaussian family and
identity link function to examine the relationship between pixel
resolution, vegetation class, and their interaction on the response
variable (the loss of vegetation area within a given class). The model
formula was as follows:

\begin{equation}\phantomsection\label{eq-gam}{
VegetationLoss \sim  Resolution \times VegetationClass
}\end{equation}

@@

In this study, drone flights were conducted at two different altitudes
(12 and 120 m) to construct the neural network model. At the sites where
the flights at both altitudes overlapped, reflectance was compared.
Overall there was a good agreement between the two altitudes (RMSE:
0.027 ; Figure~\ref{fig-CompareRef}).

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicate the model derived likelihood of the
selected class for every pixel. The low-altitude flight conducted in
Gafanha, Portugal, represented the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with green an red
macroalgae mixed with a seagrass meadow. There was also diatoms forming
biofilms on bare sediment surface. Although the seagrass was solely
composed of a single specie, \emph{Nanozostera noltei}, various colors
of this specie could be observed from dark green (corresponding to
healthy leaves) to brown (when leaves are senescent or have an altered
pigment composition). Regardless of the variation of color, the class
Magnoliopsida was accurately predicted by the model (F1 score of 0.96 at
that site).

@@

Among the high altitude flights, the one acquired over the inner part of
Ria de Aveiro coastal lagoon covered the largest area with approximately
1.5 km² (Figure~\ref{fig-Boat}). This site was dominated by seagrass and
red macroalgae. The classification provided consistent results, with a
patchy seagrass meadow mixed with red macroalgae on the eastern part of
the site. As shown in the zoom (Figure~\ref{fig-Boat}), the edges of the
meadow were mixed with green macroalgae (\emph{Ulva sp.}), which the
model agreed with (F1 score of 0.89 for green algae, 0.97 for seagrass
and 0.98 for red algae).

@@

\caption{\label{fig-Boat}RGB orthomosaic (Top) and Prediction (Bottom)
of the flight made in the inner part of Ria de Aveiro Lagoon, Portugal.
The total extent of this flight was about 1.5 km² with a resolution of
80 mm per pixel. Background colors indicate intertidal area (Light
Green), land area (Light Grey) and water (Light Blue). Each cover an
area equivalent to a 10 m Sentinel-2 pixel size.}

@@

The flight over L'Epine in Noirmoutier Island, France
(Figure~\ref{fig-Dike} A) was conducted near a dike, which crossed the
northern part of the site from West to East. Alongside this dike, Fucale
brown macroalgae (\emph{Fucus spp.}, \emph{Ascophyllum nodosum}) were
attached to sparse rocks and stranded green algae (\emph{Ulva spp.})
could be observed and this was correctly reproduced by the prediction
(Figure~\ref{fig-Dike} B). This site was characterized by a high mixture
between green macroalgae and seagrass but these two classes were
correctly discriminated by the classifier (F1 score of 0.97 and 0.98
respectively).

@@

The spectral bands at 444, 717 and 842 nm of the Micasense camera did
not provide as important information to discriminate any of the
vegetation classes. The band at 531 nm was the only important predictor
for the classifier to accurately predict Chlorophyceae. In fact, at this
wavelength, the Chlorophyceae spectra showed the highest reflectance
among all vegetation classes. The bands at 531 and 740 nm were the most
important predictors for Phaeophyceae, corresponding to the lowest
reflectance among all classes. Bands at 475 and 560 nm were the most
important predictors for Bacillariophyceae and Rhodophyceae,
respectively. Four predictors, ranging from the green (560 nm) to the
RedEdge (705 nm) bands were important to accurately predict
Magnoliopsida.

@@

The Generalized Linear Model (GLM) analysis showed a significant effect
of pixel resolution on the mapped vegetation area, indicating a positive
relationship between resolution and area loss
(\(\beta = 0.067, p < 0.001\)). This suggests that as pixel resolution
increases, the mapped area of each vegetation type decreases.

>>>>

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing can be limited by too coarse spatial
and/or spectral resolution, making it difficult to discriminate seagrass
from other macrophytes in highly heterogenous meadows. Drone (unmanned
aerial vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor similar to that onboard
Sentinel-2, for mapping intertidal macrophytes at low tide (i.e.~during
a period of emersion) and effectively discriminating between seagrass
and green macroalgae. Nine drone flights were conducted at two different
altitudes (12 m and 120 m) across heterogeneous intertidal European
habitats in France and Portugal, providing multispectral reflectance
observation at very high spatial resolution (8 mm and 80 mm
respectively). Taking advantage of their extremly high spatial
resolution; the low altitude flights were used to train a Neural Network
classifier to discrimintate five taxonomic classes of intertidal
vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae),
Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic
Bacillariophyceae (Diatoms), and validated using concomitant field
measurements. Classification of drone imagery resulted in an overall
accuracy of 94\% across all sites and images, covering a total area of
467 000 m². The model exhibited an accuracy of 96.4\% in identifying
seagrass. In particular , seagrass and green algae can be discriminated.
The very high spatial resolution of the drone data made it possible to
assess the influence of spatial resolution on the classification
outputs, showing a limited loss in seagrass detection up to about 10m.
Altogether, out findings suggest that the Multi-Spectral Instrument
(MSI) onboard Sentinel-2 offers a relevant trade-off between its spatial
and spectral resolution, thus offering promising perspectives for
satellite remote sensing of intertidal biodiversity over lager scales.
\end{abstract}

@@

The importance of seagrass meadows and the variety of ecosystem services
they provide have led to the enhancement of both global and regional
programs to monitor Essential Oceanic Variable (EOVs) such as seagrass
composition \citep{Miloslavich2018}, as well as Essential Biodiversity
Variable (EBVs) such as seagrass taxonomic diversity, species
distribution, population abundance, and phenology \citep{Pereira2013}.
Traditionally, indicators of seagrass status have been quantified using
\emph{in situ} measurements. However, the acquisition of field
measurements in intertidal zones is notoriously challenging. Intertidal
seagrass meadows are only exposed during low tide and can be situated in
difficult-to-reach mudflats, potentially leading to inaccurate and
limited estimations with conventional sampling techniques
\citep{nijland2019}. Satellite observations have been proven effective
in complementing \emph{in situ} sampling, allowing for near real-time
and consistent retrieval of seagrass EOVs and EBVs over extensive
meadows
\citep{Zoffoli2021, xu2021, Traganos2018, coffer2023, davies2024intertidal, davies2024sentinel}.

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by too
coarse spatial resolution (\textgreater100 m) to accurately map patchy
seagrass meadows. Missions with a high spatial resolution such as
Sentinel-2 (10 m) or Landsat8/9 (30 m) can be limited by low spectral
resolution. The limited number of spectral bands challenges accurate
discrimination of seagrass from other co-existing macrophytes. In
particular, Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition
\citep{ralph2002, Douay2022}, resulting in a similar spectral signature
in terms of reflectance, especially in the visible range
\citep{Davies2023, bannari2022}. Recently, using advanced
machine-learning algorithms trained with a large hyperspectral library
of more than 300 field reflectance spectra, \citep{Davies2023}
demonstrated that it was possible to discriminate Magnoliopsida from
Chlorophyceae using reflectance spectra at Sentinel-2 's spectral
resolution. However the application of this approach to satellite RS
remains to be validated. Moreover patches of green algae can develop at
small spatial scales that are not observable using Sentinel-2 and/or
Landsat-8/9 images \citep{tuya2013}, especially during the initial stage
of a green tide.

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management
\citep{adade2021, casella2020, angnuureng2022} to mapping species
distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue}.
However, when applied to coastal habitat mapping, previous case studies
were mostly limited to a low number of flights over a single study site,
restricting the generalizability of their application over wider
geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
The present study aimed at analyzing the potential of multispectral
drone RS to map intertidal macrophytes over a diverse biogeographical
range, with a particular focus on discriminating Magnoliopsida and
Chlorophyceae (Seagrass and Green Algae, respectively). Nine drone
flights were performed over soft-bottom intertidal areas along the
Atlantic coastlines of two European countries (France and Portugal),
covering a wide range of habitats, from monospecific seagrass meadows to
meadows mixed with green, or red macroalgae. A deep learning algorithm
was trained and validated for macrophyte discrimination, emphasizing
applicability across diverse sites without a loss of prediction
accuracy. The classification maps obtained at a very high spatial
resolution with the drone were spatially degraded to satellite
resolutions, making it possible to assess the effect of spatial
resolution on classification accuracy, and provide insights for coastal
habitat mapping using satellite remote sensing.

@@

Seven study sites distributed between France and Portugal were selected
for their extensive intertidal seagrass beds. Two sites were located in
the Gulf of Morbihan, France (Figure~\ref{fig-map} A : 47.5791°N,
2.8018°W). This gulf covers an area of 115 km² and is only connected to
the sea through a 900 m wide channel. A total of 53 small islands are
scattered across the gulf leading to 250 km of shorelines. Patchy
seagrass meadows can be found on many of these islands. One of the sites
within the gulf was on one its islands (Arz) and the other was located
further south on a mainland beach area (Duer). The Gulf of Morbihan is a
Natura 2000 site and a Regional Protected Area due to its rich
biodiversity, including its seagrass meadows, and is also classified as
a RAMSAR site, which highlights its significance as a wetland of
international importance. Two other sites were located in Bourgneuf Bay,
France (Figure~\ref{fig-map} B : 46.9849°N, 2.1488°W) which is a 340 km²
semi-enclosed macrotidal bay, protected from waves by Noirmoutier
Island. Bourgneuf bay hosts a large intertidal seagrass meadow of about
6 km² \citep{ZOFFOLI2020112020}. Within this meadow, the sites observed
by drones (L'Epine and Barbatre) contained monospecific beds of
\emph{Nanozostera noltei} (dwarf eelgrass, syn. \emph{Zostera noltei})
with very little mixing with other macrophytes. Bourgneuf Bay is also
part of the Natura 2000 network and serves as a RAMSAR site due to its
critical habitat for migratory bird species and its extensive seagrass
meadows \citep{Zoffoli2022}. Three sites were surveyed in the Ria de
Aveiro Coastal Lagoon in Portugal (Figure~\ref{fig-map} C : 40.6887°N,
8.6810°W). The extent of this lagoon is \textasciitilde83 km² (at low
tide) with many narrow channels, large salt marshes and many mudflats
that uncover at low tide \citep{sousa2017blue}. It is connected to the
open sea through a single channel, with a tidal lag between the North
and the South of the lagoon. The southernmost site (Gafanha) is a
mudflat located in the Mira channel (one of the four main channels of
the lagoon) whereas the two other sites (Mataducos and Marinha
Lanzarote) were situated in the middle of the lagoon and only accessible
by boat. These Portuguese sites are characterized by a more diverse
intertidal vegetation, where patches of seagrass intermingle with red,
brown, and green macroalgae. The Aveiro Lagoon, like the other study
areas, is a Natura 2000 site and a RAMSAR wetland, recognized for its
rich mosaic of habitats and importance for biodiversity, including
migratory bird species and intertidal vegetation.

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m² ,which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation : Bacillariophyceae
(unicellular benthic diatoms forming biofilms at the sediment surface
during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf
eelgrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red
macroalgae) (Figure~\ref{fig-vegetation}). Only homogeneous vegetation
patches extending over several meters were selected as ground control
points. Pictures of each quadrat were uploaded online to the open-portal
Global Biodiversity Information Facility (GBIF) platform
\citep{BedeGbif}. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to misclassification reduction in photo
interpretations.

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft} was used to process images to obtain
multispectral orthomosaics of each flight. The process for
orthomosaicking was identical for every flight. First, key tying points
were detected inside of each image and between overlapping images in
order to obtain a sparse point cloud. This cloud was cleaned using a
reprojection accuracy metric in order to remove noisy points. A dense
point cloud was then produced using a structure from motion algorithm. A
surface interpolation of this dense point cloud was made to obtain a
digital surface model (DSM), used to reconstruct the multispectral
ortho-image \citep{nebel2020review}. Low altitude drone flights produced
ortho-images with a very high spatial resolution (8 mm per pixel),
making it efficient to visually distinguish between the various types of
vegetation. High altitude flights allowed to cover larger areas and
produced images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

A dataset containing photo-interpreted drone reflectance pixels was
built to train a Neural Network model. The training pixels were
categorized into seven different classes, representing the various
habitats encountered at the different study sites: Sediment, Water,
Chlorophyceae, Magnoliopsida, Bacillariophyceae, Phaeophyceae and
Rhodophyceae. Only data from the low-altitude flights
(Table~\ref{tbl-flights}) were used for training because their 8 mm
spatial resolution allowed to avoid spectral sub-pixel mixing and to
accurately identify vegetation classes. More than 418,000 pixels at 8 mm
resolution from the 3 training flights were used to train the model
(Table~\ref{tbl-validationPX}). Twenty one variables were used by the
model as predictors: the ten raw spectral bands of the Micasense RedEdge
Dual MX multispectral camera (ranging from 444 nm to 840 nm), the same
ten spectral bands standardized using a min/max transformation
(Equation~\ref{eq-std} ; \citep{Cao2017}) and the Normalized difference
vegetation index (NDVI, Equation~\ref{eq-ndvi}). Standardisation of
spectral bands is commonly used to eliminate the scaling differences
between spectra and to limit the effect of biomass on the spectra shape
\citep{Douay2022, Davies2023}.

@@

A neural network classification model was built using the fastai
workflow \citep{howard2018fastai}. This model was composed of 2 hidden
layers and have a total of 26 054 trainable parameters. Parameters have
been fine tuned using 12 epoch to minimize the error rate. This model
has been called DISCOV, standing for Drone Intertidal Substrat
Classification Of Vegetation.

@@

The workflow of this study revolves around two distinct flight heights
(12 and 120 m, Figure~\ref{fig-workflow}) where ensuring consistency
between reflectances at both heights is crucial. This comparison was
conducted at sites where low and high altitude flights overlapped. The
low altitude flights were resampled to the same spatial resolution and
grid as the high flights using a median resampling method. Reflectance
values were then extracted, and a scatterplot was generated and the Root
Mean Square Error (RMSE) was computed to compare the difference bewteen
the raw and standardised reflectance.

The classification model was applied to all flights at both 12 and 120 m
of altitude. \emph{In situ} information on georeferenced class type and
percent cover, acquired over homogeneous vegetation patches at the same
time as drone flights was used to assess the model accuracy. These
images were used to construct a validation dataset indicating the
presence or absence of each class. Additionally to the quadrat-based
validation dataset, polygons of each class were photo interpreted in
order to increase the number of pixels of the validation dataset. A
total of 536,000 pixels were used to validate the Neural Network
classifier. The sites with the lowest and highest number of validation
data were Gafanha Low (17316 pixels) and Marinha Lanzarote (159713
pixels), respectively. A confusion matrix, along with precision metrics
such as global accuracy, sensitivity, specificity, F1 score, and Kappa
coefficient, were generated for each site. All validation matrices were
then aggregated to create an overall matrix .

@@

Variable Importance Plots (VIP) serve as a method to identify which
predictors are important for predicting a specific class. Out of the 21
predictors utilized in this study, Variable Importance was computed only
for the raw and standardized values of the 10 spectral bands captured by
the MicaSense camera. This is achieved by repeatedly predicting the same
dataset while randomly shuffling one predictor at a time. The benchmark
score obtained after each iteration is then compared to the benchmark
score obtained without shuffling any variables. The greater the
difference between these two benchmark values, the more important the
variable is for the model \citep{WEI2015399}.

@@

We used a Generalized Linear Model (GLM) with a Beta distribution to
examine the relationship between pixel resolution, vegetation class, and
their interaction on the loss of vegetation. The loss of vegetation was
modelled as function of the interaction between pixel resolution and
vegetation class (Bacillariophyceae, Phaeophyceae, Magnoliopsida,
Chlorophyceae and Rhodophyceae). Sample vs fitted residuals and
quartile-quartile graphics were assessed visually, to ensure assumption
of the models used were met

@@

In this study, drone flights were conducted at two different altitudes
(12 and 120 m) to construct the neural network model. At the sites where
the flights at both altitudes overlapped, reflectance was compared.
Overall there was a good agreement between the two altitudes (RMSE :
0.027 ; Figure~\ref{fig-CompareRef}).

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the model derived probability of the
selected class for every pixel. The low-altitude flight conducted in
Gafanha, Portugal, represented the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with green and
red macroalgae mixed with a seagrass meadow. There was also diatoms
forming biofilms on bare sediment surface. Although the seagrass was
solely composed of a single specie, \emph{Nanozostera noltei}, various
colors of this specie could be observed from dark green (corresponding
to healthy leaves) to brown (when leaves are senescent or have an
altered pigment composition). Regardless of the variation of color, the
class Magnoliopsida was accurately predicted by the model (F1 score of
0.96 at that site).

@@

Among the high altitude flights, the one acquired over the inner part of
Ria de Aveiro coastal lagoon covered the largest area with approximately
1.5 km² (Figure~\ref{fig-Boat}). Teh vegetation present at the site was
dominated by seagrass and red macroalgae. The classification provided
consistent results, with a patchy seagrass meadow mixed with red
macroalgae on the eastern part of the site. As shown in the zoom
(Figure~\ref{fig-Boat}), the edges of the meadow were mixed with green
macroalgae (\emph{Ulva sp.}), which the model agreed with (F1 score of
0.89 for green algae, 0.97 for seagrass and 0.98 for red algae).

@@

\caption{\label{fig-Boat}RGB orthomosaic (Top) and Prediction (Bottom)
of the flight made in the inner part of Ria de Aveiro Lagoon, Portugal.
The total extent of this flight was about 1.5 km² with a resolution of
80 mm per pixel. Background colors indicate intertidal area (Light
Green), land area (Light Grey) and water (Light Blue). The zoom inserts
cover an area equivalent to the size of a 10-meter Sentinel-2 pixel.}

@@

The flight over L'Epine in Noirmoutier Island, France
(Figure~\ref{fig-Dike} A) was conducted near a dike, which crossed the
northern part of the site from West to East. Alongside this dike, Fucale
brown macroalgae (\emph{Fucus spp.}, \emph{Ascophyllum nodosum}) were
attached to sparse rocks, and stranded green algae (\emph{Ulva spp.})
could be observed, which was correctly reproduced by the prediction
(Figure~\ref{fig-Dike} B). This site was characterized by a high mixture
between green macroalgae and seagrass but these two classes were
correctly discriminated by the classifier (F1 score of 0.97 and 0.98
respectively).

@@

The spectral bands at 444, 717 and 842 nm of the Micasense camera did
not provide important information to discriminate any of the vegetation
classes. The band at 531 nm was the most important predictor by far for
the classifier to accurately predict Chlorophyceae. In fact, at this
wavelength, the Chlorophyceae spectra showed the highest reflectance
among all vegetation classes (Figure~\ref{fig-vegetation} F). The bands
at 531 and 740 nm were the most important predictors for Phaeophyceae,
corresponding to the lowest reflectance among all classes. Bands at 475
and 560 nm were the most important predictors for Bacillariophyceae and
Rhodophyceae, respectively. Four predictors, ranging from the green (560
nm) to the RedEdge (705 nm) bands were important to accurately predict
Magnoliopsida.

@@
